Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-05-31 main.tex
Added in data processing figure.
Added perturbation figure.
Update for CI build corrections.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} seems to be the first \todo{Check if a Japanese
researcher was first} to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly predicted. In his
model, he recognized that feedback of the vehicle roll angle and angular rate
was essential to stabilize the vehicle. It was not until the mid 1980s that
Ruijs and Pacjecka successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for
favorable control at all speeds. This roll motion feedback is the simplest
controller that can stabilize a single track vehicle above a minimum speed when
not concerned with wobble instabilities, which are not prominent in bicycles at
low speeds. Ruijs and Pacejka's work was not particularly concerned with low
speed stability and their vehicle was fully automatic, no human rider was
involved. Many more automatically balanced single track vehicles have been
demonstrated over the last 40 years, but none have demonstrated that increasing
low speed stability can assist in balancing and possibly reduce single actor
falls. Most of these robotic bicycles and motorcycles designers did not intend
for a human rider to also control the stabilized vehicle. Nevertheless, an
automatically stabilized bicycle can be controlled by a human rider if the
motor controlled steer torque and the rider applied steer torque act on the
steer in parallel. The effect of this automatic control gives the ability to
change the dynamics, up to some limits, of the human controlled plant.In our
prior study, we demonstrate reduced motion during
distractions~\cite{Alizadehsaravi2023} due to the balance assist but
researchers with a similar vehicle recently showed rider disatisfction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. Whipple even showed that roll motion feedback can stabilize the
system, remarkably in the early years of developments automatic control. The
bicycle model is ideal for showing the effect of a roll motion feedback driven
steer motor on the dynamics. The linear model can be described by the fourth
order state space equations:

@@

\begin{figure}
  \includegraphics{example-image-a}
  \caption{Force nd torque plts of a pterubation of 80N per rope in the
  counterclockwise direction (from Marten's thesis, pg 26)}
\end{figure}

\begin{figure}
  \includegraphics{example-image-a}
  \caption{Thesis Fig 3.13 on pg 27}
\end{figure}

@@

>>>>

In the early years of developments automatic control\citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize the a
bicycle. Automatic roll stabilization of single track vehicles began in earnest
after predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
the first \todo{Check if a Japanese researcher was first} to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the mid 1980s that Ruijs and Pacjecka
successfully demonstrated an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
propose how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when not
concerned with wobble instabilities, which are not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycles and motorcycles designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant.In our prior study, we
demonstrate reduced motion during distractions~\cite{Alizadehsaravi2023} due to
the balance assist but researchers with a similar vehicle recently showed rider
disatisfction with the stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear model can be
described by the fourth order state space equations:

@@

\begin{figure}
  \includegraphics[width=\textwidth]{figures/perturbation_0.png}
  \caption{Force nd torque plts of a pterubation of 80N per rope in the
  counterclockwise direction (from Marten's thesis, pg 26)}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/data-processing.png}
  \caption{Thesis Fig 3.13 on pg 27}
\end{figure}
